---
title: "Pruning as Confound: How Maintenance Processes Obscure Functional Diversity in Self-Organising Networks"
subtitle: "Weight Pruning, Metric Confounds, and the Conditions for Functional Differentiation"
author:
  - name: Michael Bean
    affiliation: Independent Researcher
format:
  html:
    toc: true
    code-fold: true
    fig-width: 7
    fig-height: 5
bibliography: references.bib
---

```{r setup, include=FALSE}
source("_setup.R")
fig_dir <- file.path(getwd(), "figures")
meta_data <- file.path(dirname(dirname(getwd())), "data")  # Points to repo data/ directory

bound_colours <- c("hard"="#E63946", "tanh"="#457B9D", "oja"="#E9C46A")
bound_labels <- c("hard"="Hard clip", "tanh"="Tanh", "oja"="Oja")
```

## Abstract

Papers 1--4 of this programme studied Hebbian self-organisation under a
standard network maintenance regime: edges with absolute weight below a
pruning threshold ($10^{-4}$) are removed each step. This maintenance
operation interacts disordinally with weight bounding strategy. Under
pruning, hard-clipped substrates produce 3.9 assemblages while tanh-bounded
substrates produce 0.2. Disabling pruning reverses the ranking: tanh
produces 7.0, hard clip 5.6. The interaction amplifies with network scale
(15$\times$15: $\Delta=4.2$; 30$\times$30: $\Delta=14.5$) and is
independent of edge-cap mechanism (identical results at caps of 5k, 10k,
20k, 50k, and unlimited). The centroid cosine distance metric is
indistinguishable from a spatial null ($p = 1.0$, 1000 permutations). A
deconfounded response protocol using uniform stimuli reveals response rank
$5.6$ (participation ratio $2.2$; robust across SVD thresholds from $0.1\%$
to $10\%$). No-pruning replications show pruning actively destroyed
functional differentiation: rank rises from $1.0$ (pruned) to $5.6$
(no-prune) even without metaplasticity. The hard clip is qualitatively
irreducible---no finite steepness recovers its behaviour---functioning as a
limit condition^[In Simondon's [-@simondon2020individuation] terminology: a
*condition limite* that constrains the space of possible individuations,
distinct from both a *moule* (mould, which imposes form) and a *germe
structural* (structural germ, which templates form through propagation).]
that scaffolds diversity emerging from Hebbian dynamics on spatially
structured substrates.


## Introduction

The Dark Forest research programme studies how self-organising Hebbian
networks produce individuation---the emergence of distinct, persistent
assemblages from homogeneous initial conditions. Papers 1--3 established
that individuation occurs readily under hard weight clipping
($w = \text{clip}(w, -w_\text{max}, w_\text{max})$), with
locality-dependent coexistence, self-sealing, and structural persistence.
Paper 4 introduced metaplasticity (per-node learning rates with homeostatic
adjustment) and demonstrated that local homeostatic targets stabilise
functional differentiation.

Throughout the programme, a standard maintenance operation---pruning edges
below $|w| < 10^{-4}$---was applied without systematic variation. This
paper demonstrates that pruning is not neutral housekeeping. It is an active
structural intervention that interacts with weight bounding strategy in ways
that distort the programme's central comparisons. Recent work in deep
network compression has established that pruning disproportionately harms
underrepresented classes [@hooker2019compressed; @tran2022pruning], with
clinical consequences in medical imaging [@holste2023pruning] and
measurable subgroup disparities of up to 90\% [@lin2022fairgrape]. Our
findings extend this pattern to self-organising substrates where no
external training signal exists: the weak connections destroyed by pruning
are not noise obscuring pre-existing structure but the differentiation
mechanism itself.

The pruning threshold functions as an unintended *mould* in Simondon's
sense [@simondon2020individuation]: an external constraint that imposes
form by selectively eliminating the weak exploratory connections through
which soft-bounded substrates organise. Removing this mould reveals that
soft bounds produce more assemblages than hard bounds, that the programme's
primary differentiation metric (CCD) is spatially confounded, and that
pruning was suppressing genuine functional diversity. The hard weight clip,
by contrast, functions as a *limit condition*: a boundary that constrains
the dynamics without imposing specific form, creating the structural
scaffold on which functional diversity develops.


## Results

### The hard clip is qualitatively irreducible

We parameterised the bounding function as $w_\text{max} \cdot \tanh(k
\cdot w / w_\text{max})$, sweeping steepness $k$ from 1 (standard tanh) to
50 (near-hard) with pruning disabled ($n = 20$ seeds, 100 sessions,
locality $= 0.10$).

```{r fig-steepness, fig.cap="Fig. 1: Non-monotonic steepness result. Both the softest bound (k=1) and the hard clip produce multiple assemblages. Intermediate steepness (k=2--50) produces monopoly.", fig.width=7, fig.height=5}
d7b <- read_experiment("exp7b_steepness_noprune")
d2b <- read_experiment("exp2b_baseline_noprune")

if (!is.null(d7b) && !is.null(d2b)) {
  d7b_final <- d7b |>
    filter(session == max(session)) |>
    group_by(steepness) |>
    summarise(asm = mean(n_assemblages), se = sd(n_assemblages)/sqrt(n()), .groups = "drop")

  hard_ref <- d2b |>
    filter(session == max(session), bound_type == "hard") |>
    summarise(asm = mean(n_assemblages), se = sd(n_assemblages)/sqrt(n()))

  ggplot(d7b_final, aes(x = steepness, y = asm)) +
    geom_line(linewidth = 0.8, colour = "#457B9D") +
    geom_point(size = 2.5, colour = "#457B9D") +
    geom_errorbar(aes(ymin = asm - se, ymax = asm + se), width = 0.08, colour = "#457B9D") +
    geom_hline(yintercept = hard_ref$asm, linetype = "dashed", colour = "#E63946", linewidth = 0.6) +
    annotate("text", x = 30, y = hard_ref$asm + 0.3, label = "Hard clip", colour = "#E63946", size = 3.5) +
    scale_x_log10(breaks = c(1, 2, 3, 5, 10, 20, 50)) +
    labs(x = "Steepness k (log scale)", y = "Assemblages (no pruning)") +
    theme(panel.grid.minor = element_blank())
}
```

The result is non-monotonic (Fig. 1). At $k = 1$ (tanh): 6.0 assemblages
with $\bar{w} = 0.005$ and low activity ($0.053$). At $k \geq 1.5$:
weights saturate ($\bar{w} \approx 1.0$), activity rises to $0.57$, and
competitive exclusion drives the system toward monopoly ($1.0$--$1.8$
assemblages). The hard clip ($4.9$ assemblages) is not recovered at any
finite steepness. The sharp transition at $k = 1.5$ corresponds to the
onset of weight saturation (0\% to 86\% of edges above $0.95 w_\text{max}$)
and a tenfold increase in mean activation.


### The pruning confound: a disordinal interaction

We tested the pruning--bound-type interaction with a $2 \times 3$ factorial
design: pruning on (threshold $10^{-4}$) or off (threshold $10^{-8}$),
crossed with three weight bounding strategies (hard clip, tanh, oja;
sigmoid omitted as it is identical to tanh to $\geq 4$ decimal places).
Each cell: $n = 30$ seeds, 200 sessions, locality $= 0.10$.

```{r fig-factorial, fig.cap="Fig. 2: Disordinal interaction between pruning and bound type (n=30 per cell). With pruning, hard clip dominates. Without pruning, tanh and oja produce more assemblages.", fig.width=8, fig.height=5}
dF <- read_experiment("expF_factorial")
if (!is.null(dF)) {
  dF_final <- dF |>
    filter(session == max(session), bound_type != "sigmoid") |>
    mutate(prune_label = ifelse(prune_on, "Pruned (1e-4)", "No pruning")) |>
    group_by(bound_type, prune_label) |>
    summarise(asm = mean(n_assemblages), se = sd(n_assemblages)/sqrt(n()), .groups = "drop")

  ggplot(dF_final, aes(x = bound_type, y = asm, fill = prune_label)) +
    geom_col(position = position_dodge(0.8), width = 0.7, alpha = 0.85) +
    geom_errorbar(aes(ymin = asm - se, ymax = asm + se),
                  position = position_dodge(0.8), width = 0.2) +
    scale_fill_manual(values = c("Pruned (1e-4)" = "#264653", "No pruning" = "#2A9D8F")) +
    scale_x_discrete(labels = bound_labels) +
    labs(x = "Weight bound type", y = "Assemblages at session 200",
         fill = "Pruning regime") +
    theme(legend.position = "bottom")
}
```

The interaction is disordinal (Fig. 2). Under pruning, hard clip produces
$3.9 \pm 2.1$ assemblages while tanh produces $0.2 \pm 0.4$. Without
pruning, tanh produces $7.0 \pm 2.2$ while hard clip produces
$5.6 \pm 3.2$. The ranking reverses.

A threshold sweep identifies the critical region between $10^{-6}$
(tanh $= 13.9$ assemblages) and $10^{-5}$ (tanh $= 0.7$), corresponding
to the equilibrium weight magnitude for soft-bounded edges. Hard bounds are
insensitive ($3.4$--$4.0$ across the entire range).

**Edge cap independence.** The no-pruning condition uses a rank-based edge
cap to prevent unbounded memory growth. A sensitivity sweep over cap values
(5,000; 10,000; 20,000; 50,000; unlimited) shows identical assemblage
counts at every value (hard: $5.6$; tanh: $6.9$; oja: $8.2$). Edge count
self-stabilises at $\sim 1{,}350$ on the $20 \times 20$ grid---well below
any cap---because the Hebbian dynamics reach a natural equilibrium between
encounter-driven edge creation and decay-driven edge loss. The cap is never
activated.

**Grid size generality.** The interaction persists and amplifies across
network scales:

| Grid | Hard $\Delta$ | Tanh $\Delta$ | Oja $\Delta$ |
|---|---|---|---|
| $15 \times 15$ | $+0.0$ | $+4.2$ | $+4.2$ |
| $20 \times 20$ | $+1.7$ | $+6.7$ | $+7.0$ |
| $30 \times 30$ | $+7.8$ | $+14.5$ | $+15.2$ |

The pruning effect on soft bounds scales with network size, consistent with
the mechanism: larger networks sustain more weak edges in the critical
$10^{-6}$--$10^{-5}$ range.


### Pruning suppressed genuine functional differentiation

No-pruning replications of Paper 4's perturbation-response experiment
reveal that pruning actively destroyed functional diversity.

```{r fig-rank, fig.cap="Fig. 3: Response rank with and without pruning. Pruning destroyed assemblages in the local hard condition entirely. Without pruning, both baseline and local hard reach rank 5.6.", fig.width=8, fig.height=5}
d17_orig <- tryCatch(read_parquet(file.path(meta_data, "exp17_response_matrix", "session_metrics.parquet")), error = function(e) NULL)
d17_np <- tryCatch(read_parquet(file.path(meta_data, "exp17np_response_noprune", "session_metrics.parquet")), error = function(e) NULL)

if (!is.null(d17_orig) && !is.null(d17_np)) {
  make_summary <- function(d, label) {
    d |> filter(n_assemblages >= 2) |>
      group_by(condition) |>
      summarise(rank = mean(response_rank), se = sd(response_rank)/sqrt(n()),
                n = n(), .groups = "drop") |>
      mutate(pruning = label)
  }

  combined <- bind_rows(
    make_summary(d17_orig, "Pruned"),
    make_summary(d17_np, "No pruning")
  ) |>
    filter(condition != "global_target") |>
    mutate(condition = recode(condition,
      "baseline" = "Baseline\n(no meta)",
      "local_hard" = "Local hard\n(meta=0.05)",
      "local_target" = "Local hard\n(meta=0.05)",
      "local_soft" = "Local soft\n(meta=0.05)"))

  ggplot(combined, aes(x = condition, y = rank, fill = pruning)) +
    geom_col(position = position_dodge(0.8), width = 0.7, alpha = 0.85) +
    geom_errorbar(aes(ymin = rank - se, ymax = rank + se),
                  position = position_dodge(0.8), width = 0.2) +
    scale_fill_manual(values = c("Pruned" = "#264653", "No pruning" = "#E76F51")) +
    labs(x = NULL, y = "Response rank (effective dimensions)",
         fill = "Pruning regime") +
    theme(legend.position = "bottom")
}
```

Under pruning, response rank is $1.0$ across all conditions (flat across
all SVD thresholds from $0.1\%$ to $10\%$). Without pruning, rank rises to
$5.6$ for hard bounds and $5.5$ for tanh (Fig. 3). The rank estimate is
robust across SVD threshold choices:

| Threshold | Hard (no-prune) | Tanh (no-prune) | Oja (no-prune) |
|---|---|---|---|
| $0.1\%$ | $5.9$ | $5.8$ | $6.4$ |
| $1\%$ (default) | $5.6$ | $5.5$ | $5.8$ |
| $5\%$ | $4.8$ | $4.8$ | $5.1$ |
| $10\%$ | $4.1$ | $4.2$ | $4.5$ |

The participation ratio---a threshold-free measure of effective
dimensionality computed as $(\sum \sigma_i^2)^2 / \sum \sigma_i^4$ over
the singular values $\sigma_i$---gives $2.2$ (hard), $2.1$ (tanh), and
$2.2$ (oja) under Gaussian probes, rising to $3.2$, $3.1$, and $3.3$
under uniform (spatially deconfounded) probes.

The baseline condition (no metaplasticity) reaches the same rank as the
metaplastic condition. Hard clipping alone produces functional diversity.
Metaplasticity adds $\eta$ divergence (below) and increases assemblage
count from $\sim 6$ to $10+$, but is not necessary for functional
differentiation per se.


### Eta differentiation requires absorbing boundaries

Per-assemblage learning rate ($\eta$) variance was measured under soft
$\eta$ bounds ($n = 50$ seeds, 300 sessions). Of 50 seeds, 6 produced
$\geq 2$ assemblages; the remaining 44 converged to monopoly (single
assemblage). The failure mode is stochastic: monopoly and multi-assemblage
seeds show identical $\eta$ statistics ($\bar{\eta} = 0.098$,
$\sigma_\eta = 0.019$) and edge counts ($\sim 1{,}310$), and the
qualifying seed count fluctuates between 5 and 8 across sessions with no
trend, confirming that coexistence under hard weight + soft $\eta$ bounds
is a rare but genuine dynamical regime.

For the 6 qualifying seeds:

| Statistic | Value |
|---|---|
| Mean $F$-ratio (between/within) | $4.91$ (SD $= 0.70$) |
| Bootstrap 95\% CI (BCa, 10K) | $[4.40, 5.42]$ |
| $t(5) = 13.77$, $p$ (one-tailed) | $0.000018$ |
| Cohen's $d$ | $5.62$ |

The $F$-ratio represents the ratio of between-assemblage $\eta$ variance to
within-assemblage $\eta$ variance, computed separately within each seed and
then aggregated via one-sample $t$-test against $F = 1$ across seeds. This
avoids pseudoreplication from pooling assemblages across seeds. All soft
weight bound conditions show $F < 1$ (tanh: $0.86$; oja: $0.76$). The
absorbing weight boundary creates the structural asymmetry that the
metaplastic instability requires.


### CCD is spatially confounded

Centroid cosine distance was tested against a spatial null: weight values
were randomly permuted across edges (preserving the edge topology but
randomising which edges carry which weights), assemblages were re-detected
on the permuted weights using the same adaptive threshold (75th percentile),
and CCD was recomputed. This was repeated 1,000 times per seed ($n = 20$
seeds per condition). No condition reached $p < 0.05$.

A deconfounded response protocol (Exp R2) revealed that Gaussian probes
inflated JS divergence by $0.27$ (spatial overlap between probe position
and assemblage location) while deflating response rank from $3.1$--$3.3$
(uniform probes) to $2.0$--$2.2$ (Gaussian). The spatial structure of
Gaussian probes created correlated activation across assemblages,
collapsing independent functional dimensions.


## Three Regimes of Individuation

The experiments support three nested dynamical regimes, distinguished by
their structural and functional properties. These are not static categories
but three phases of the same Hebbian process, observed under different
boundary conditions:

| Regime | Assemblages | Response rank | $\eta$ $F$-ratio | Example |
|---|---|---|---|---|
| **Topological** | 6--8 | 1--2 | $<1$ | Soft bounds, no pruning |
| **Structural** | 5--6 | 5.6 | --- | Hard bounds, no pruning, no meta |
| **Metaplastic** | 10+ | 5.6 | 4.9 | Hard bounds, no pruning, local homeostasis |

**Topological individuation**: structure through connectivity patterns with
negligible weight magnitude ($\bar{w} \approx 0.005$). Detection is
threshold-sensitive. Assemblages respond to stimuli with low functional
diversity (rank $1$--$2$).

**Structural individuation**: the absorbing weight boundary creates
saturated edges that resist competitive displacement, producing a rigid
scaffold. Detection is threshold-robust. Functional diversity (rank $5.6$)
emerges without metaplasticity.

**Metaplastic individuation**: local homeostasis drives $\eta$ divergence
between assemblages ($F = 4.9$), increasing assemblage count and adding
parametric differentiation---different learning dynamics across assemblages,
not just different weight structures. Requires both the hard clip boundary
and a local homeostatic mechanism.

These are nested: each regime subsumes the previous. Structural
individuation requires the hard clip. Metaplastic individuation requires
both the hard clip and local homeostasis.


## Discussion

### Pruning as unintended mould

The pruning threshold functions as a second absorbing boundary---one that
was never theorised as part of the experimental mechanism. For hard-bounded
edges (equilibrium $|w| \approx 1.0$), pruning at $10^{-4}$ removes only
dead connections. For soft-bounded edges (equilibrium
$|w| \approx 10^{-5.5}$), pruning destroys the substrate before
individuation can occur. In Simondon's [-@simondon2020individuation]
terminology, pruning is an unintended *mould*: an external constraint that
imposes form by eliminating structures that fall below its threshold,
regardless of their functional contribution.

The disordinal interaction is direct evidence that the system's
individuating capacity depends on boundary conditions in a way that cannot
be predicted from either boundary alone. The pruning confound has
implications beyond this programme: any computational model that removes
weak connections may be destroying exactly the exploratory structure that
enables functional diversity. This parallels concerns in the synaptic
pruning literature, where pruning is typically modeled as constructive
(improving generalization by removing noise) rather than destructive
[@chechik1998synaptic]. Scholl, Rule, and Hennig [-@scholl2021information]
showed formally that information-theoretic pruning criteria reduce
generative diversity in deep Boltzmann machines, even when improving
efficiency on surviving representations. Our work extends this in two
directions: we characterise the *geometric* mechanism (response rank
collapse, participation ratio reduction^[This parallels the $1/n$
eigenspectrum observed in biological visual cortex by
@stringer2019high --- healthy representational spaces maintain a specific
spectral profile that pruning disrupts.]), and we demonstrate that the
interaction is *disordinal* --- not merely that pruning hurts some
conditions more than others, but that it reverses the ranking of approaches
entirely. The vulnerability of weakly-represented structure to pruning
echoes a broader pattern: Granovetter [-@granovetter1973strength] showed
that weak ties carry disproportionate novel information in social networks;
analogously, low-magnitude connections in self-organising networks encode
precisely the exploratory structure that enables functional diversity. In
the neural network compression literature, Frankle and Carlin
[-@frankle2019lottery] showed that dense networks contain sparse
"winning ticket" subnetworks that match full-network performance; our
weak inter-assemblage connections are the complementary phenomenon ---
"losing tickets" that carry no individual predictive value but whose
collective presence scaffolds representational diversity. In biological
development, silent synapses --- connections that carry no current but
serve structural roles during critical periods --- provide a direct
analog: they appear inactive by electrophysiological measures but are
essential for circuit maturation.

### The hard clip as limit condition

The non-monotonic steepness result demonstrates that the hard clip is not
the infinite-steepness limit of soft saturation. It is qualitatively
different: a *limit condition* in Simondon's [-@simondon2020individuation]
sense. A limit condition constrains the space of possible individuations
without imposing specific form---it tells the system where it cannot go, not
what it must become. The hard clip at $\pm w_\text{max}$ creates an
absorbing state that edges can reach but not leave, producing a bimodal
weight distribution that functions as a structural scaffold for functional
diversity.

This is distinct from a *mould* (which imposes external form, as pruning
does). The hard clip does not propagate from a local singularity as a seed
crystal would---it constrains uniformly. But by constraining, it creates the asymmetric
conditions under which the metaplastic dynamics can generate
between-assemblage $\eta$ divergence.

### Programme implications

The core narrative of the Dark Forest programme holds, but the mechanism
is clarified. Papers 1--3 established individuation, coexistence, and
self-sealing under hard bounds. Paper 4 demonstrated metaplastic
differentiation. This paper reveals that pruning distorted the quantitative
comparisons, that CCD was spatially confounded, and that functional
differentiation is substantially stronger than originally reported (response
rank $5.6$ vs $2.4$). The revised claim: local homeostatic metaplasticity
stabilises and amplifies functional differentiation that Hebbian dynamics on
spatially structured substrates spontaneously produce, provided the
maintenance regime does not destroy the evidence.


## Methods

### Weight bounding strategies

Four strategies were compared: hard clip
($w = \text{clip}(w, -w_\text{max}, w_\text{max})$), tanh saturation
($w = w_\text{max} \tanh(w/w_\text{max})$), sigmoid saturation, and Oja
normalisation. Sigmoid $\equiv$ tanh across all experiments to $\geq 4$
decimal places and is omitted from analysis.

### Pruning conditions

Standard pruning: edges with $|w| < 10^{-4}$ removed each step.
No-pruning: threshold $10^{-8}$ (effectively disabled). To prevent
unbounded edge accumulation, a rank-based cap removes the weakest edges
when total count exceeds a limit (default 20,000). A sensitivity analysis
(caps at 5,000; 10,000; 20,000; 50,000; unlimited) confirms the cap never
activates: edge count self-stabilises at $\sim 1{,}350$ on the
$20 \times 20$ grid.

### Response rank

After 200 training sessions, $K = 10$ structured input patterns (Gaussian
bumps at random grid positions) are injected without further learning. Each
assemblage's mean response is recorded, producing a $K \times A$ response
matrix. Effective rank: number of singular values exceeding a threshold
percentage of the largest. We report rank at five thresholds ($0.1\%$,
$0.5\%$, $1\%$, $5\%$, $10\%$) and the participation ratio
$(\sum \sigma_i^2)^2 / \sum \sigma_i^4$ as a threshold-free alternative.
Deconfounded probes use spatially uniform random stimuli
$U[-0.3, 0.3]$ per node.

### CCD null model

For each seed, weight values are randomly permuted across edges (preserving
edge topology but randomising weight assignment). Assemblages are
re-detected on the permuted weights using the same adaptive threshold
(75th percentile of absolute weight). CCD is recomputed. This is repeated
1,000 times. The $p$-value is the proportion of null CCD values $\geq$ the
observed CCD.

### Statistical methods

Effect sizes: Cohen's $d$ with bootstrap CIs (10,000 BCa). $\eta$
variance: per-seed $F$-ratios (between-assemblage / within-assemblage
variance) aggregated via one-sample $t$-test against $F = 1$ across seeds,
avoiding pseudoreplication from pooling assemblages across independent
simulations. The 44/50 seeds converging to monopoly are reported as a
failure-mode analysis rather than excluded silently.


## References
