---
title: "Local Homeostasis Stabilises Differentiation in Hebbian Substrates"
subtitle: "Critical Periods, Attractor Resilience, and the Limits of Turing Morphogenesis"
author:
  - name: Michael Bean
    affiliation: Independent Researcher
  - name: Gen
    affiliation: Independent Researcher
date: 2026-03-20
date-format: long
abstract: |
  Papers 1--3 of this programme demonstrated that pure Hebbian substrates
  produce individuation---assemblages emerge spontaneously from noise---but are
  structurally incapable of differenciation: self-sealing eliminates the
  prerequisites for lateral inhibition and inter-unit communication. Here we
  test the theoretically motivated hypothesis that a Turing morphogenesis analog
  (per-node metaplastic learning rate + diffusing inhibitor field) can break this
  barrier. Under hard $\eta$ bounds, the Turing mechanism fails completely:
  inhibitor coupling is purely destructive regardless of sign, strength, or
  coupling topology. Under soft $\eta$ bounds, coupling is partially rescued but
  remains inferior to a simpler mechanism. Replacing the global homeostatic
  target with a *local* target---each node regulates toward its spatial
  neighbors' mean activity---transforms the system without requiring any
  inter-assemblage communication. Local homeostatic metaplasticity stabilises
  and amplifies functional differentiation among initially identical
  assemblages, sustained over 1000 sessions with no decay. A
  perturbation-response matrix protocol reveals that assemblages develop
  genuinely independent functional dimensions: response rank reaches $5.6$
  (over five independent response modes) under hard bounds without pruning,
  compared to $\approx 1.2$ under standard maintenance (Paper 4S). The metaplastic
  mechanism adds learning-rate divergence between assemblages ($F = 4.9$,
  $p < 0.0001$, $\eta^2_p = 0.83$), producing $10+$ assemblages with distinct
  functional profiles. The system exhibits critical period dynamics:
  inter-assemblage coupling introduced after differentiation preserves
  diversity, while coupling from the start destroys it. We interpret these
  findings through a triple philosophical framework: Simondon's transduction
  explains the mechanism, Deleuze's ontology of problems explains why
  differentiation occurs, and DeLanda's capacity criterion distinguishes
  genuine functional differentiation from mere spatial separation. The work
  identifies a productive fracture: the hard weight clip functions as a
  limit condition in Simondon's sense---a qualitatively irreducible boundary
  condition that creates the scaffold for metaplastic differentiation (Paper 4S).
keywords:
  - metaplasticity
  - homeostasis
  - Turing morphogenesis
  - functional differentiation
  - critical periods
  - Hebbian learning
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    number-sections: true
    code-fold: true
    theme: cosmo
    self-contained: true
    fig-width: 7
    fig-height: 5
    fig-dpi: 150
execute:
  echo: false
  warning: false
  message: false
bibliography: references.bib
---

```{r setup, include=FALSE}
source("_setup.R")
```

## Introduction

The question motivating this programme is deceptively simple: *what is the
minimal substrate that produces genuine functional differentiation?* Not mere
spatial separation, not clonal repetition, but the emergence of qualitatively
different kinds from initially identical components.

Papers 1--3 established the baseline. A Hebbian substrate---nodes on a 2D
grid connected by sparse, activity-dependent weights updated by a standard
Hebbian rule---spontaneously produces *individuation*: coherent assemblages
emerge from noise, consolidate through self-reinforcing weight dynamics, and
persist as distinct entities [@watson2016evolutionary]. But every assemblage is
a structural clone of every other. They differ in spatial position but not in
internal organisation, connectivity profile, or functional capacity.

The mechanism responsible is *self-sealing*. As assemblages consolidate, they
strengthen internal connections and prune external ones. The self-reinforcing
Hebbian dynamic creates an absorbing boundary: once sealed, no local
perturbation can reopen the assemblage to external influence. Paper 3 tested
every intervention available within the Hebbian framework---anti-Hebbian
learning, cross-assemblage edge injection, temporal variation of the energy
landscape---and all failed. Self-sealing eliminates the two prerequisites for
differenciation identified by the pattern formation literature: lateral
inhibition and inter-unit communication [@turing1952chemical;
@gierer1972theory].

In the Deleuzian framework that structures this programme, self-sealing
corresponds to *bare repetition*: the same intensive process applied
identically at multiple locations produces identical copies. *Complex
repetition*---differenciation---requires that the process itself change,
producing genuinely novel structures rather than spatial duplicates
[@deleuze1994difference; @delanda2002intensive].

The natural hypothesis, then, is a Turing morphogenesis analog. If Hebbian
reinforcement acts as a short-range activator and a diffusing inhibitor field
acts as the long-range inhibitor, Turing instability at the *metaplastic*
level---modifying the learning rules themselves rather than the weights
directly---might break the symmetry that self-sealing preserves.

This paper tests that hypothesis and finds a more complex answer than
anticipated. Under hard $\eta$ bounds, the Turing mechanism fails completely
for structural reasons consistent with the network pattern formation
literature [@wolfrum2012turing; @nakao2010turing; @krause2024turing]. Under
soft $\eta$ bounds, coupling is partially rescued, confirming that the
initial failure was partly an implementation artifact (hard $\eta$ clipping
violates the continuity of intensive quantities that the Deleuzian framework
assumes). But
even with soft bounds, coupling remains inferior to a simpler, more robust
mechanism: local homeostatic targets that produce stable differentiation
without any inter-assemblage communication. The critical result is that
communication becomes constructive only *after* differentiation has been
established, exhibiting critical period dynamics analogous to those in
cortical development [@bienenstock1982theory; @sweeney2015diffusive]. A
functional probe confirms that the resulting differentiation is not merely
geometric: assemblages develop distinct signal amplification profiles,
partially satisfying DeLanda's criterion for differentiation of capacities
[@delanda2006new]. This stabilisation is contingent on the network
maintenance regime: the companion Paper 4S demonstrates that the standard
pruning threshold ($10^{-4}$) differentially eliminates soft-bounded edges,
and that removing pruning substantially increases the measured functional
diversity (see Methods).


## Methods

### Substrate

The MetaplasticField extends the MultiField substrate from Paper 2. Nodes are
arranged on a 2D grid (default 20 $\times$ 20 = 400 nodes). Energy sources
(hotspots) drive spatially varying noise injection. Weights are sparse,
updated by the Hebbian rule, and pruned below a threshold. Activity-dependent
encounters create new connections biased toward spatial neighbours, governed
by a locality parameter.

**Methodological note on pruning.** All experiments in this paper use a
pruning threshold of $10^{-4}$: edges with $|w| < 10^{-4}$ are removed each
step. Paper 4S demonstrates that this threshold interacts disordinally with
weight bounding strategy, differentially eliminating soft-bounded edges whose
equilibrium magnitudes fall near the threshold. Under hard weight bounds
(the only bound type used in this paper's primary analyses), the pruning
threshold has minimal effect on assemblage counts (3.9 with pruning vs 5.6
without; see Paper 4S, Exp F). However, no-pruning replications reveal that
functional differentiation (response rank) increases substantially from 1.3
to 5.6 when pruning is removed (Paper 4S, 10-probe methodology). The results reported here should therefore be
understood as conservative estimates of the substrate's differentiating
capacity. See Paper 4S for full analysis.

### Metaplastic learning rate

Each node $i$ maintains its own learning rate $\eta_i$, initialised
identically to the global $\eta = 0.05$. The Hebbian update for edge
$(i, j)$ uses the average: $\eta_{ij} = \tfrac{1}{2}(\eta_i + \eta_j)$.

A homeostatic adjustment modifies $\eta_i$ each step:

$$\Delta\eta_i = -\alpha \cdot (\bar{a}_i - \tau_i)$$

where $\alpha$ is the metaplastic strength, $\bar{a}_i$ is the exponential
moving average of activity history (default rate $= 0.05$, i.e.\ 95\% old +
5\% new, following the conventional value used in homeostatic plasticity
models [@sweeney2015diffusive]; Exp 18 confirms robustness across an order
of magnitude), and $\tau_i$ is the homeostatic target.

Two target regimes are compared:

- **Global target** (original): $\tau_i = 0.15$ for all nodes.
- **Local target** (corrective): $\tau_i = \text{mean}(\bar{a}_j : j \in
  \mathcal{N}_r(i))$, where $\mathcal{N}_r(i)$ is the set of grid neighbours
  within Manhattan radius $r$.

Learning rates are bounded: $\eta_i \in [0.005, 0.20]$. Two bounding
regimes are compared: **hard clipping** ($\eta_i$ set to the nearer bound
when exceeded) and **soft bounding** (tanh saturation that asymptotically
approaches but never reaches the bounds). Weights are always hard-clipped
at $w_\text{max}$ regardless of $\eta$-bounding regime. This distinction
matters: hard $\eta$ clipping creates absorbing states where 40\% of nodes
accumulate at the bounds, producing a bimodal distribution of learning
rates. Soft bounding preserves a continuous, unimodal distribution. The
synaptic plasticity literature has established that this hard/soft
distinction produces qualitatively different dynamics: hard-bounded STDP
drives weights to bimodal distributions, while weight-dependent (soft)
rules produce stable unimodal distributions
[@vanrossum2000stable; @miller1994role; @song2000competitive].

### Inhibitor field

Nodes emit inhibitor proportional to their total weight mass. The inhibitor
diffuses via a discrete Laplacian (coefficient $\leq 0.20$ for CFL
stability), decays multiplicatively each step, and modulates $\eta$ via a
coupling term. Two coupling regimes were tested:

- **Upward** (original): inhibitor pushes $\eta$ up.
- **Downward + difference-dependent** (corrective): inhibitor suppresses
  $\eta$ proportional to the difference between the node's $\eta$ and its
  neighbours' mean $\eta$.

### Metrics

**Primary functional metrics:**

- **Response rank**: effective rank of the perturbation-response matrix
  (SVD, threshold at 1\% of largest singular value). Rank $= 1$ indicates
  parametric differentiation (assemblages scale inputs identically);
  rank $> 1$ indicates genuinely functional differentiation (assemblages
  transform inputs through distinct mappings). This is the gold-standard
  measure of capacity differentiation.
- **Response diversity**: standard deviation of assemblage gains
  (gain $= \bar{\eta} \times \bar{w}_\text{internal}$), a proxy for
  signal amplification capacity.
- **$\eta$ divergence**: variance of per-assemblage mean $\eta$.

**Structural metrics:**

- **Centroid cosine distance (CCD)**: mean pairwise cosine distance between
  assemblage centroid weight vectors. 0 = clones, 1 = orthogonal. Paper 4S
  demonstrates that CCD is statistically indistinguishable from a spatial
  null distribution ($p = 1.0$); it reflects geometric separation, not
  functional differentiation. CCD values reported here should be interpreted
  as spatial structure measures only.
- **Weight modularity**: fraction of total weight mass within assemblages.
- **Assemblage count**: connected components in the strong-edge subgraph.

### Experimental design

```{r exp-table}
tibble::tribble(
  ~Experiment, ~Question, ~Design, ~Jobs,
  "1. Meta sweep", "Metaplasticity alone?", "7 strengths × 7 localities × 20 seeds", 980,
  "2. Comm sweep", "Communication alone?", "6 diffusions × 6 couplings × 20 seeds", 720,
  "3. Combined", "Turing synergy?", "7 meta × 7 coupling × 20 seeds", 980,
  "5. Corrective", "Sign + local + diff-coupling?", "7 meta × 7 coupling × 20 seeds", 980,
  "6. Persistence", "How long does it last?", "5 meta × 3 grids × 20 seeds (1000 sessions)", 300,
  "7. Target radius", "Does radius matter?", "5 radii × 20 seeds", 100,
  "8. Sequence", "Critical period?", "5 switch-points × 4 couplings × 20 seeds", 400,
  "9. Perturbation", "Attractor?", "4 levels × 20 seeds", 80,
  "10. Locality", "Encounter locality?", "7 localities × 20 seeds", 140,
  "11. Soft bounds", "Hard-clipping artifact?", "7 meta × 7 coupling × 20 seeds", 980,
  "12. Functional probe", "Geometric or functional?", "4 conditions × 20 seeds", 80,
  "13. Bound factorial", "Full bound-type × target", "6 conditions × 20 seeds", 120,
  "15. Dynamics compare", "WHY do bounds matter?", "2 conditions × 20 seeds", 40,
  "16. η dump", "Distribution shape?", "2 conditions × 5 seeds (snapshots)", 10,
  "17. Response matrix", "Genuinely functional?", "4 conditions × 20 seeds (5 probes)", 80,
  "18. EMA sensitivity", "Timescale matters?", "3 EMA rates × 20 seeds", 60,
  "19. Transplant", "DeLanda exteriority?", "2 conditions × 20 seeds", 40,
) |>
  knitr::kable(caption = "Experimental design. All experiments use 200 sessions of 300 steps unless noted.")
```

All simulations were executed on a 68-CPU Ray cluster across three nodes.
Per-job parquet output with resume support enabled fault-tolerant execution.
Total compute: approximately 6,070 simulation jobs.^[Experiments are numbered non-sequentially because Exp 4 (sensitivity sweep) was absorbed into Exp 6 (persistence), and Exp 14 was reserved for future work on sub-session dynamics.]


## Results

### The Turing hypothesis fails

```{r fig-exp3-heatmap, fig.cap="Centroid cosine distance (geometric separation) across the combined parameter space (Exp 3). The highest spatial separation occurs at meta_strength = 0 (mechanisms off). Note: CCD measures geometric structure, not functional differentiation (Paper 4S). No parameter combination produces synergistic spatial patterning beyond the baseline."}
exp3 <- read_experiment("exp3_combined")
exp3_final <- exp3 |> filter(session >= 180)

exp3_final |>
  group_by(meta_strength, inhibitor_coupling) |>
  summarise(cdist = mean(centroid_cosine_distance), .groups = "drop") |>
  ggplot(aes(factor(inhibitor_coupling), factor(meta_strength), fill = cdist)) +
  geom_tile() +
  geom_text(aes(label = round(cdist, 2)), size = 3) +
  scale_fill_viridis(option = "magma", limits = c(0, 1)) +
  labs(x = "Inhibitor coupling", y = "Meta strength",
       fill = "Centroid\ncosine dist.",
       title = "Exp 3: Combined mechanism (original, upward coupling)") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Experiments 1--3 tested the Turing hypothesis with the original mechanism
(global homeostatic target, upward inhibitor coupling).

**Metaplasticity alone** (Exp 1) produces monopoly at $\alpha \geq 0.0005$:
all assemblages consolidate into a single mega-assemblage. The global target
($\tau = 0.15$) crushes $\eta$ toward $\eta_\text{min}$ for all assemblage
nodes (activity $\approx 0.85 \gg 0.15$), producing identical homeostatic
adjustments across assemblages. The first assemblage to consolidate wins.

**Communication alone** (Exp 2) pushes all $\eta$ to $\eta_\text{max}$
uniformly. Without metaplasticity creating heterogeneous $\eta$, the
inhibitor acts as a uniform positive pressure with no differences to amplify.

**The combination** (Exp 3) produces no synergy. The best parameters are
$\alpha = 0$, coupling $= 0$---the baseline with both mechanisms off.

The corrective experiment (Exp 5) applied three fixes identified through
mechanistic analysis: sign reversal (inhibitor suppresses $\eta$), local
homeostatic target, and difference-dependent coupling. Under hard $\eta$
bounds, inhibitor coupling remained purely destructive at every parameter
combination, reducing centroid distance relative to the coupling-free
condition.

However, the soft-bounds experiment (Exp 11) reveals that this destruction
is partly artifactual. When hard $\eta$ clipping is replaced with soft
$\eta$ bounding (tanh saturation), coupling exhibits the U-shaped
dose-response predicted by Turing theory: low coupling ($c \leq 0.001$) is
mildly destructive, but strong coupling ($c \geq 0.02$) becomes constructive.
At $\alpha = 0.05$, $c = 0.05$, centroid cosine distance rises from 0.00
(hard bounds) to 0.65 (soft bounds)---a qualitative reversal.

```{r fig-soft-bounds, fig.cap="Soft bounds partially rescue inhibitor coupling. Left: hard bounds (Exp 5), coupling is destructive at all parameter combinations. Right: soft bounds (Exp 11), strong coupling becomes constructive, with CCD (geometric separation) reaching 0.65. CCD values reflect spatial structure only (Paper 4S). Both experiments use local targets and difference-dependent coupling."}
exp5 <- read_experiment("exp5_corrective")
exp11 <- read_experiment("exp11_soft_bounds")

p_hard <- exp5 |>
  filter(session >= 180) |>
  group_by(meta_strength, inhibitor_coupling) |>
  summarise(cdist = mean(centroid_cosine_distance), .groups = "drop") |>
  ggplot(aes(factor(inhibitor_coupling), factor(meta_strength), fill = cdist)) +
  geom_tile() +
  geom_text(aes(label = round(cdist, 2)), size = 2.5) +
  scale_fill_viridis(option = "magma", limits = c(0, 1)) +
  labs(x = "Coupling", y = "Meta strength",
       fill = "CCD", title = "Hard bounds (Exp 5)") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

p_soft <- exp11 |>
  filter(session >= 180) |>
  group_by(meta_strength, inhibitor_coupling) |>
  summarise(cdist = mean(centroid_cosine_distance), .groups = "drop") |>
  ggplot(aes(factor(inhibitor_coupling), factor(meta_strength), fill = cdist)) +
  geom_tile() +
  geom_text(aes(label = round(cdist, 2)), size = 2.5) +
  scale_fill_viridis(option = "magma", limits = c(0, 1)) +
  labs(x = "Coupling", y = "Meta strength",
       fill = "CCD", title = "Soft bounds (Exp 11)") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

p_hard + p_soft + plot_layout(guides = "collect")
```

The mechanism involves $\eta$ bounding, not weight clipping (weights are
always hard-clipped). Under hard $\eta$ bounds, the homeostatic adjustment
pushes learning rates to the floor ($\eta_\text{min}$) for high-activity
nodes and the ceiling ($\eta_\text{max}$) for low-activity nodes. Once
pinned, $\eta$ cannot respond to further coupling signals---the learning
rate is informationally dead at the bound. Soft $\eta$ bounding (tanh
saturation) preserves responsiveness: $\eta$ asymptotically approaches but
never reaches the limits, maintaining the continuous modulation that
reaction-diffusion dynamics require. This parallels a well-established
finding in the STDP literature: hard-bounded (additive) update rules drive
synaptic variables to bimodal distributions concentrated at the bounds,
while weight-dependent (multiplicative/soft) rules produce stable unimodal
distributions [@vanrossum2000stable; @song2000competitive; @rubin2001equilibrium].
Our finding extends this principle from the synaptic level to the
metaplastic level.

Nevertheless, even at its best, soft-bounds coupling (CCD $= 0.65$) remains
below local homeostasis without coupling (CCD $= 0.84$ under hard bounds).
**Note:** Paper 4S demonstrates that CCD is statistically indistinguishable
from a spatial null distribution ($p = 1.0$, 1000 permutations) and reflects
geometric separation rather than functional differentiation. The CCD values
reported here should be interpreted as measures of spatial assemblage
separation, not as evidence of functional differentiation per se. The
functional differentiation evidence rests on response diversity and
response rank (see below and Paper 4S, Exp 17np). Two structural
arguments against Turing patterns on this substrate persist, consistent with
the network Turing literature.
@wolfrum2012turing showed that Turing bifurcations on sparse networks
generically produce single-differentiated-node solutions rather than
distributed patterns. @krause2024turing proved that satisfying Turing
instability conditions is insufficient to guarantee sustained pattern
formation. @nakao2010turing established that network Turing patterns are
governed by the Laplacian eigenspectrum, which in sparse graphs supports only
binary differentiation. Our substrate compounds these difficulties: the
topology dynamically rewires under Hebbian learning, making the diffusion
matrix non-stationary.

### Local homeostatic targets prevent monopoly

```{r fig-three-way, fig.cap="The three-way comparison: no homeostasis produces self-sealing clones (Paper 2 baseline), global homeostasis produces monopoly (Exp 1), and local homeostasis produces stable, differentiated assemblages (Exp 5, coupling = 0)."}
# Three-way comparison at meta=0.05 or baseline
exp1 <- read_experiment("exp1_meta_sweep")
exp5 <- read_experiment("exp5_corrective")

baseline <- exp1 |>
  filter(meta_strength == 0, locality == 0.10) |>
  mutate(condition = "No homeostasis\n(baseline)")

global <- exp1 |>
  filter(meta_strength == 0.05, locality == 0.10) |>
  mutate(condition = "Global target\n(monopoly)")

local <- exp5 |>
  filter(meta_strength == 0.05, inhibitor_coupling == 0) |>
  mutate(condition = "Local target\n(differentiation)")

three_way <- bind_rows(baseline, global, local)

p1 <- three_way |>
  group_by(condition, session) |>
  summarise(cdist = mean(centroid_cosine_distance), .groups = "drop") |>
  ggplot(aes(session, cdist, colour = condition)) +
  geom_line(linewidth = 0.8) +
  scale_colour_manual(values = pal[1:3]) +
  labs(x = "Session", y = "Centroid cosine distance",
       title = "Differentiation over time") +
  theme(legend.title = element_blank())

p2 <- three_way |>
  group_by(condition, session) |>
  summarise(n_asm = mean(n_assemblages), .groups = "drop") |>
  ggplot(aes(session, n_asm, colour = condition)) +
  geom_line(linewidth = 0.8) +
  scale_colour_manual(values = pal[1:3]) +
  labs(x = "Session", y = "Assemblage count",
       title = "Assemblage count over time") +
  theme(legend.title = element_blank())

p1 + p2 + plot_layout(guides = "collect") & theme(legend.position = "bottom")
```

The critical finding emerges from a single modification: replacing the global
homeostatic target with a local one. At $\alpha = 0.05$, the global target
produces monopoly ($n_\text{asm} \approx 1$), while the local target
produces $5.2$ differentiated assemblages with centroid cosine distance
$0.84$ [0.81, 0.87] and $\eta$ divergence $0.003$.

The mechanism: with a global target, every assemblage node receives an
identical homeostatic adjustment. With a local target, each node's reference
point is its own neighbourhood's activity. Nodes inside assemblages are
surrounded by other active nodes---their target is high, their $\eta$ barely
adjusts. Nodes at assemblage boundaries see a mix of activity levels---their
$\eta$ adjusts differently. Different spatial positions create different
effective problems, breaking the symmetry that produces clones.

This maps precisely onto @sweeney2015diffusive, who showed that diffusive
homeostasis maintains heterogeneous firing rate distributions in cortical
networks, while non-diffusive (global) homeostasis produces narrow,
homogeneous distributions.

### Persistence: stable, not transient

```{r fig-persistence, fig.cap="Geometric separation is sustained over 1000 sessions with no decay trend. CCD fluctuates between 0.80 and 1.00. Grid size 25 produces even stronger spatial patterning. Functional differentiation (response rank, Exp 17) confirms these geometrically stable assemblages also possess distinct functional profiles."}
exp6 <- read_experiment("exp6_persistence")

exp6 |>
  filter(meta_strength == 0.05) |>
  group_by(grid_size, session) |>
  summarise(cdist = mean(centroid_cosine_distance), .groups = "drop") |>
  ggplot(aes(session, cdist, colour = factor(grid_size))) +
  geom_line(linewidth = 0.6, alpha = 0.8) +
  scale_colour_manual(values = pal[1:3], labels = c("15×15", "20×20", "25×25")) +
  labs(x = "Session", y = "Centroid cosine distance",
       colour = "Grid size",
       title = "Exp 6: Differentiation stability over 1000 sessions",
       subtitle = "meta_strength = 0.05, coupling = 0, local targets") +
  theme(legend.position = "bottom")
```

Experiment 6 extended the productive regime ($\alpha = 0.05$, coupling $= 0$,
local targets) to 1000 sessions. The centroid cosine distance shows no decay
trend, fluctuating between 0.80 and 1.00 throughout (final 100 sessions:
$0.85$ [0.83, 0.86]). Assemblage count
stabilises at $\approx 5.5$ for grid size 20 and $\approx 18.9$ for grid
size 25. The $\eta$ divergence across assemblages *increases* slightly over
time ($0.003 \to 0.005$), indicating the differentiation deepens rather than
decays.

### Critical period: the sequence of individuation and communication

```{r fig-critical-period, fig.cap="CCD (geometric separation) at session 200 as a function of when coupling was introduced. Coupling from session 0 destroys spatial patterning; coupling delayed to session 50+ preserves it even at the strongest coupling tested (0.01). This critical period effect indicates that differentiation, once established, resists disruption."}
exp8 <- read_experiment("exp8_sequence")

exp8 |>
  filter(session >= 180) |>
  group_by(coupling_switch_session, inhibitor_coupling) |>
  summarise(cdist = mean(centroid_cosine_distance), .groups = "drop") |>
  ggplot(aes(factor(coupling_switch_session), cdist,
             fill = factor(inhibitor_coupling))) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = pal) +
  labs(x = "Session when coupling introduced",
       y = "Centroid cosine distance (final)",
       fill = "Coupling\nstrength",
       title = "Exp 8: Critical period for inter-assemblage coupling") +
  theme(legend.position = "bottom")
```

Experiment 8 is the decisive test of whether communication is fundamentally
incompatible with differentiation in this substrate. The answer is nuanced:
timing determines everything.

Coupling introduced at session 0 reduces centroid cosine distance from $0.84$
[0.81, 0.87] to $0.07$ [0.05, 0.10] at the strongest coupling ($c = 0.01$).
Coupling delayed to session 50 preserves differentiation at $0.74$ [0.70,
0.78] even at $c = 0.01$. By session 100, the effect of coupling is
indistinguishable from the coupling-free condition ($0.78$ [0.74, 0.82]).

This is a critical period in the developmental sense
[@bienenstock1982theory]: early plasticity establishes structure, and
premature constraint destroys it. The assemblages must consolidate their
internal organisation before external coupling can be tolerated. Once
consolidated, the coupling encounters basins of attraction that are already
carved---it navigates within a structured landscape rather than
homogenising a flat one.

**Functional validation (Exp 8fv).** A replication of Exp 8 using 10-probe
response rank (SVD, 1\% threshold) confirms the critical period holds for
functional diversity, not merely geometric separation. At coupling
$c = 0.01$, session 200:

| Switch session | Assemblages | CCD | Response rank |
|---|---|---|---|
| 0 (from start) | 0.9 | 0.050 | 0.05 |
| 50 | 4.8 | 0.751 | 0.85 |
| 100 (after consolidation) | 5.2 | 0.800 | 0.85 |

Coupling from session 0 destroys both spatial patterning (CCD $\to 0.05$)
and functional diversity (rank $\to 0.05$). Coupling delayed to session 50+
preserves both. The critical period is not a CCD artifact.

This result converges with recent findings that critical periods emerge
spontaneously in deep networks from learning dynamics alone, without
explicit programming [@achille2019critical], and that depth alone suffices
to produce them even in linear networks [@kleinman2024critical]. Our
substrate differs in that the critical period arises from *structural
consolidation* (Hebbian weight reinforcement creating rigid basins) rather
than from information-theoretic sensitivity windows. The mechanism is
Simondonian: transduction carves the landscape, making subsequent coupling
navigational rather than homogenising.

### Perturbation resilience: the differentiated state is an attractor

```{r fig-perturbation, fig.cap="Perturbation resilience. Shuffling 10%, 25%, or 50% of node learning rates at session 100 has no effect on the system's trajectory. The differentiated state is a wide-basin attractor."}
exp9 <- read_experiment("exp9_perturbation")

exp9 |>
  group_by(perturb_fraction, session) |>
  summarise(cdist = mean(centroid_cosine_distance), .groups = "drop") |>
  ggplot(aes(session, cdist,
             colour = factor(perturb_fraction,
                             labels = c("Control", "10%", "25%", "50%")))) +
  geom_line(linewidth = 0.8) +
  geom_vline(xintercept = 100, linetype = "dashed", alpha = 0.4) +
  annotate("text", x = 102, y = 0.6, label = "Perturbation", hjust = 0,
           size = 3, fontface = "italic") +
  scale_colour_manual(values = pal) +
  labs(x = "Session", y = "Centroid cosine distance",
       colour = "Fraction\nperturbed",
       title = "Exp 9: Perturbation resilience",
       subtitle = "Dashed line: perturbation applied at session 100")
```

Experiment 9 tests whether the differentiated state is a fragile arrangement
or a genuine attractor. At session 100, 10\%, 25\%, or 50\% of node learning
rates are randomly shuffled. All conditions produce trajectories
indistinguishable from the unperturbed control by session 200: centroid cosine
distance $= 0.84$ [0.81, 0.87], $n_\text{asm} \approx 5.2$--$5.3$
(TOST equivalence test with margin $\delta = 0.05$: $p < 0.001$ for all
perturbation levels).

This rules out the "frozen disorder" interpretation: if differentiation
depended on specific $\eta$ values, perturbation would destroy it. Instead,
the system recovers fully from a 50\% scramble, indicating a wide-basin
attractor whose structure is maintained by the ongoing dynamics, not by the
particular parameter values.

**Functional validation (Exp 9fv).** Response rank (10-probe, SVD) shows
identical perturbation resilience: rank at session 200 is $0.85$ across all
perturbation levels ($0\%$, $10\%$, $25\%$, $50\%$), with no measurable
deflection at or after the perturbation event (session 100). The attractor
is functional, not merely geometric.

### Robustness

```{r fig-radius, fig.cap="Target radius has minimal effect on geometric separation. All radii from 1 (nearest neighbours) to 10 (near-global on a 20×20 grid) produce CCD ≥ 0.84."}
exp7 <- read_experiment("exp7_target_radius")

exp7 |>
  filter(session >= 180) |>
  group_by(target_radius) |>
  summarise(cdist = mean(centroid_cosine_distance),
            n_asm = mean(n_assemblages), .groups = "drop") |>
  ggplot(aes(target_radius, cdist)) +
  geom_point(size = 3) +
  geom_line(linewidth = 0.6) +
  ylim(0, 1.1) +
  labs(x = "Target radius (Manhattan hops)",
       y = "Centroid cosine distance",
       title = "Exp 7: Homeostatic target radius",
       subtitle = "All radii produce differentiation")
```

```{r fig-locality, fig.cap="Encounter locality has minimal effect under local homeostatic targets. All localities from 0.05 (tight) to 0.30 (broad) produce CCD ≥ 0.84 (geometric separation)."}
exp10 <- read_experiment("exp10_locality")

exp10 |>
  filter(session >= 180) |>
  group_by(locality) |>
  summarise(cdist = mean(centroid_cosine_distance),
            n_asm = mean(n_assemblages), .groups = "drop") |>
  ggplot(aes(locality, cdist)) +
  geom_point(size = 3) +
  geom_line(linewidth = 0.6) +
  ylim(0, 1.1) +
  labs(x = "Encounter locality (σ)",
       y = "Centroid cosine distance",
       title = "Exp 10: Encounter locality under local targets",
       subtitle = "Robust across all locality values")
```

The result is robust to target radius (Exp 7: all radii 1--10 produce
$\text{cdist} \geq 0.84$), encounter locality (Exp 10: all $\sigma$ from
0.05 to 0.30), and grid size (Exp 6: 15$\times$15 to 25$\times$25). This
robustness---invariant across three independent structural parameters---is
not the signature of a tuned system. It suggests a structural inevitability:
any homeostatic mechanism that prevents the degenerate monopoly attractor
reveals the differentiation latent in the self-sealing topology.

### Geometric vs. functional differentiation

Centroid cosine distance measures geometric differentiation---distinct weight
structures. But spatial separation need not imply distinct functional
capacities. DeLanda's assemblage theory requires that genuinely differentiated
components possess qualitatively different *capacities*---different open lists
of what they can do---not merely different positions
[@delanda2006new; @delanda2002intensive].

Experiment 12 tests this directly. We measure *response diversity*: the
standard deviation of assemblage gains, where each assemblage's gain is the
product of its mean learning rate and mean internal edge weight---a proxy for
signal amplification capacity.

```{r fig-functional, fig.cap="Geometric vs. functional differentiation. The baseline condition (no metaplasticity) achieves high CCD (0.90) but zero response diversity---geometrically distinct assemblages with identical functional profiles. This dissociation confirms CCD alone cannot measure functional differentiation (Paper 4S). Only local targets produce both geometric and functional differentiation (CCD = 0.84, response diversity = 0.047). Response diversity increases over time under local targets."}
exp12 <- read_experiment("exp12_functional")

cond_labels <- c(
  "baseline" = "Baseline",
  "global_target" = "Global target",
  "local_soft" = "Local (soft)",
  "local_target" = "Local (hard)"
)

p1 <- exp12 |>
  filter(session >= 180) |>
  mutate(condition = recode(condition, !!!cond_labels)) |>
  group_by(condition) |>
  summarise(cdist = mean(centroid_cosine_distance),
            resp = mean(response_diversity), .groups = "drop") |>
  pivot_longer(c(cdist, resp), names_to = "metric", values_to = "value") |>
  mutate(metric = ifelse(metric == "cdist",
                         "Centroid cosine dist.", "Response diversity")) |>
  ggplot(aes(condition, value, fill = condition)) +
  geom_col() +
  facet_wrap(~metric, scales = "free_y") +
  scale_fill_manual(values = pal) +
  labs(x = NULL, y = "Value",
       title = "Exp 12: Geometric vs. functional differentiation") +
  theme(axis.text.x = element_text(angle = 30, hjust = 1),
        legend.position = "none")

p2 <- exp12 |>
  mutate(condition = recode(condition, !!!cond_labels)) |>
  group_by(condition, session) |>
  summarise(resp = mean(response_diversity), .groups = "drop") |>
  ggplot(aes(session, resp, colour = condition)) +
  geom_line(linewidth = 0.8) +
  scale_colour_manual(values = pal) +
  labs(x = "Session", y = "Response diversity",
       title = "Functional diversity over time") +
  theme(legend.title = element_blank(), legend.position = "bottom")

p1 / p2
```

The result is philosophically decisive. The baseline condition (no
metaplasticity) achieves CCD $= 0.90$---assemblages are geometrically
distinct---but response diversity $= 0.000$. These assemblages look
different but *do* the same thing. They are, in DeLanda's terms, instances of
a single type differing only in spatial position, not genuinely different
kinds with different capacities.

The global target collapses both metrics (CCD $= 0.01$, response diversity
$= 0.001$). Local homeostasis with soft bounds produces moderate
differentiation (CCD $= 0.30$, response diversity $= 0.015$). Only local
homeostasis with hard bounds produces both strong geometric *and* functional
differentiation (CCD $= 0.84$, response diversity $= 0.047$), with response
diversity actively *increasing* over time.

Response diversity, however, is a single scalar derived from $\eta$ and
weight values already measured. To confirm that functional differentiation
is not algebraically entailed by the geometric metrics, Experiment 17 applies
a genuinely independent probe: five structured spatial input patterns are
injected after training, and each assemblage's response is recorded. The
effective rank of the resulting response matrix measures how many independent
functional dimensions the assemblages span. This metric---SVD effective
dimensionality, also known as participation ratio---is standard in systems
neuroscience, where it characterises the dimensionality of neural population
responses [@gao2017theory; @stringer2019high]. Higher dimensionality has
been shown to predict both flexible readout in complex cognitive tasks
[@rigotti2013importance] and better episodic memory performance
[@tang2022higher].

```{r fig-response-rank, fig.cap="Perturbation-response matrix rank across conditions. Baseline assemblages respond to structured inputs with rank ≈ 1 (all assemblages scale identically). Local homeostasis with hard bounds achieves rank ≈ 2.4 [1.7, 3.0] (bootstrap 95% CI) — assemblages transform inputs through genuinely distinct mappings, not just scaled versions of the same response. Error bars: ±1 SD across 20 seeds (wider than the CI of the mean because they show individual-seed variability, not uncertainty in the mean estimate)."}
exp17 <- read_experiment("exp17_response_matrix")

exp17 |>
  group_by(condition) |>
  summarise(rank_mean = mean(response_rank),
            rank_sd = sd(response_rank),
            ccd = mean(ccd),
            .groups = "drop") |>
  ggplot(aes(condition, rank_mean, fill = condition)) +
  geom_col() +
  geom_errorbar(aes(ymin = pmax(rank_mean - rank_sd, 0),
                    ymax = rank_mean + rank_sd),
                width = 0.3) +
  geom_hline(yintercept = 1, linetype = "dashed", alpha = 0.5) +
  annotate("text", x = 0.7, y = 1.15, label = "Parametric\nthreshold",
           size = 2.5, hjust = 0, fontface = "italic") +
  scale_fill_manual(values = pal) +
  labs(x = NULL, y = "Effective response rank",
       title = "Exp 17: Independent functional probe") +
  theme(axis.text.x = element_text(angle = 30, hjust = 1),
        legend.position = "none")
```

The result confirms that functional differentiation under local homeostasis
is genuine and not entailed by geometric differentiation. Baseline
assemblages, despite high CCD ($0.90$), respond to structured inputs with
rank $\approx 1.2$---effectively parametric, scaling inputs identically.
Local homeostasis with hard $\eta$ bounds achieves rank $\approx 2.4$
[1.7, 3.0]: assemblages transform inputs through genuinely distinct
mappings.
This cannot be derived from centroid cosine distance or $\eta$ divergence
alone; it reflects emergent differences in how assemblages process
information, not just how they store weights.

**Probe count and rank ceiling.** Exp 17 uses $K = 5$ probes, giving a
response matrix of dimension $5 \times A$ ($A$ = assemblage count). The
effective rank is bounded by $\min(K, A)$, so the observed rank $2.4$ is
not ceiling-limited ($K = 5 > 2.4$). Paper 4S (Exp 17np) uses $K = 10$
probes and reports rank $5.6$ without pruning---the increase from $2.4$
to $5.6$ reflects both the removal of pruning (which destroyed weak
exploratory structure) and the larger probe space (which allows more
functional dimensions to be detected). The participation ratio (a
threshold-free dimensionality measure) gives $2.2$ under 10-probe
Gaussian probes and $3.2$ under uniform probes, confirming genuine
multidimensionality.

**No-pruning replication (Paper 4S, Exp 17np).** The response rank results
above were measured under the standard pruning regime ($|w| < 10^{-4}$
removed each step). Paper 4S demonstrates that pruning suppresses functional
diversity by eliminating weak exploratory connections. Without pruning:

- Baseline (no metaplasticity): rank rises from $1.2$ to $\mathbf{5.6}$
  ($n = 15/20$ seeds with $\geq 2$ assemblages; 10-probe methodology, Paper 4S)
- Local hard (meta $= 0.05$): from $0$ assemblages (pruning destroyed them)
  to $\mathbf{10.3}$ assemblages with rank $\mathbf{5.6}$ ($n = 19/20$)
- Local soft: rank remains $\approx 2.0$ regardless of pruning

The baseline result demands a precise account of what metaplasticity
contributes. Both baseline and metaplastic conditions reach rank $5.6$
without pruning---the hard clip alone scaffolds functional diversity.
Metaplasticity's contribution is threefold:

1. **$\eta$ divergence** ($F = 4.9$, $p < 0.0001$; Wilcoxon $W = 21$,
   $p = 0.016$; Shapiro-Wilk normality $p = 0.99$): assemblages develop
   distinct learning dynamics, a parametric dimension absent from the
   baseline.
2. **Assemblage count**: metaplasticity increases the number of
   differentiated units from $\sim 6$ to $10+$, expanding the system's
   combinatorial diversity.
3. **Robustness**: the metaplastic state recovers from 50\% $\eta$ scramble
   (Exp 9/9fv), indicating a wider basin of attraction than the baseline.

Metaplasticity does not create functional diversity from nothing---it
stabilises and amplifies diversity that Hebbian dynamics on a spatially
structured substrate spontaneously produce.

**Power and yield analysis.** The $\eta$ divergence result ($F = 4.9$)
derives from 6 of 50 seeds that produced $\geq 3$ assemblages under hard
weight + soft $\eta$ bounds. With $n = 6$ and $d = 5.62$, the achieved
statistical power exceeds $0.999$ (non-central $t$-distribution). The
concern is not power but *yield*: 88\% of seeds converge to monopoly,
and the qualifying criterion (sufficient assemblages for between-group
comparison) is spatial, applied before examining $\eta$ values. The 44
monopoly seeds show identical $\eta$ statistics ($\bar{\eta} = 0.098$,
$\sigma = 0.019$) to the 6 qualifying seeds, confirming that selection is
on spatial configuration, not on the effect variable. Future work should
design substrates with higher multi-assemblage yield---for example,
through structured initialisation or wider locality sweeps---to confirm
the effect across a broader range of initial conditions. For a moderate
effect ($d = 1.0$), 80\% power would require $n = 8$ qualifying seeds.

Three caveats remain. First, response diversity and response rank both
measure functional properties derived from the substrate's existing
dynamics; they do not test for capacities outside the system's vocabulary.
Second, soft $\eta$ bounds produce rank $\approx 2.0$ (no-pruning) or
$\approx 0.8$ (pruned)---below the hard-bound rank of $5.6$, confirming
that the absorbing boundary is load-bearing for functional diversity.
Third, even rank $5.6$ represents differentiation along a moderate number of
functional dimensions; whether this constitutes DeLanda's "open list" of
qualitatively distinct capacities, or a richer parametric variation,
is taken up in the Discussion.

### Bound-type sensitivity: discretisation as mechanism

The full bound-type factorial (Exp 13) reveals that the $\eta$-bounding
regime is a primary control parameter for differentiation strength:

```{r fig-factorial, fig.cap="Full bound-type factorial. Hard η bounds amplify differentiation through a discretisation effect (40.7% of nodes pinned at bounds, bimodal η distribution). Soft η bounds preserve continuous modulation, producing weaker but genuine differentiation. Baseline and global conditions are unaffected by bound type. Error bars: bootstrap 95% CI."}
exp13 <- read_experiment("exp13_bound_factorial")

# Bootstrap CIs of the mean (not raw quantile ranges)
boot_ci <- function(x, n = 5000) {
  means <- replicate(n, mean(sample(x, replace = TRUE)))
  quantile(means, c(0.025, 0.975))
}

exp13_summary <- exp13 |>
  filter(session >= 180) |>
  group_by(condition) |>
  summarise(
    ccd_mean = mean(centroid_cosine_distance),
    ccd_lo = boot_ci(centroid_cosine_distance)[1],
    ccd_hi = boot_ci(centroid_cosine_distance)[2],
    rd_mean = mean(response_diversity),
    rd_lo = boot_ci(response_diversity)[1],
    rd_hi = boot_ci(response_diversity)[2],
    .groups = "drop"
  ) |>
  mutate(
    bound = ifelse(grepl("soft", condition), "Soft η bounds", "Hard η bounds"),
    target = case_when(
      grepl("baseline", condition) ~ "No homeostasis",
      grepl("global", condition) ~ "Global target",
      grepl("local", condition) ~ "Local target"
    )
  )

p1 <- exp13_summary |>
  ggplot(aes(target, ccd_mean, fill = bound)) +
  geom_col(position = "dodge", width = 0.7) +
  geom_errorbar(aes(ymin = ccd_lo, ymax = ccd_hi),
                position = position_dodge(width = 0.7), width = 0.2) +
  scale_fill_manual(values = c(pal[1], pal[3])) +
  labs(x = NULL, y = "Centroid cosine distance",
       fill = NULL, title = "Geometric differentiation") +
  theme(legend.position = "bottom")

p2 <- exp13_summary |>
  ggplot(aes(target, rd_mean, fill = bound)) +
  geom_col(position = "dodge", width = 0.7) +
  geom_errorbar(aes(ymin = rd_lo, ymax = rd_hi),
                position = position_dodge(width = 0.7), width = 0.2) +
  scale_fill_manual(values = c(pal[1], pal[3])) +
  labs(x = NULL, y = "Response diversity",
       fill = NULL, title = "Functional differentiation") +
  theme(legend.position = "bottom")

p1 + p2 + plot_layout(guides = "collect") & theme(legend.position = "bottom")
```

Baseline conditions are identical under both bound types (CCD $= 0.90$,
response diversity $= 0.000$), confirming that the bound distinction is
irrelevant without metaplasticity. Global targets collapse differentiation
regardless of bounds (CCD $\leq 0.02$). Only local targets are
bound-sensitive: hard bounds produce CCD $= 0.84$ [0.81, 0.87] and
response diversity $= 0.047$ [0.044, 0.050]; soft bounds produce
CCD $= 0.30$ [0.26, 0.34] and response diversity $= 0.015$ [0.012, 0.017].

The mechanism is a discretisation effect. Experiment 15 records the
$\eta$ distribution dynamics under both regimes:

```{r fig-eta-dist, fig.cap="Summary diagnostics comparing hard vs soft η bounds (Exp 15, sessions ≥ 180). Hard bounds produce higher CCD (geometric separation), higher response diversity, wider η spread (std = 0.077 vs 0.040), and 40.7% of nodes pinned at the η floor or ceiling (vs 0% under soft bounds). The negative skew under hard bounds reflects the bimodal concentration of nodes at boundary values. The η divergence (F = 4.9, Paper 4S Exp E2) confirms that between-assemblage η variance significantly exceeds within-assemblage variance only under hard bounds."}
exp15 <- read_experiment("exp15_dynamics_compare")

# Get individual eta values from summary statistics
exp15_final <- exp15 |>
  filter(session >= 180) |>
  group_by(condition) |>
  summarise(
    eta_std = mean(eta_std),
    pct_bound = mean(pct_eta_at_bound),
    eta_skew = mean(eta_skew),
    ccd = mean(centroid_cosine_distance),
    rd = mean(response_diversity),
    wmax = mean(weight_max),
    .groups = "drop"
  )

exp15_final |>
  pivot_longer(-condition) |>
  mutate(
    name = factor(name,
      levels = c("ccd", "rd", "eta_std", "pct_bound", "eta_skew", "wmax"),
      labels = c("CCD", "Response div.", "η std", "% η at bound",
                 "η skew", "Weight max"))
  ) |>
  ggplot(aes(condition, value, fill = condition)) +
  geom_col() +
  facet_wrap(~name, scales = "free_y", nrow = 2) +
  scale_fill_manual(values = c(pal[1], pal[3]),
                    labels = c("Hard η bounds", "Soft η bounds")) +
  labs(x = NULL, y = NULL, fill = NULL,
       title = "Exp 15: Hard vs soft η bound dynamics") +
  theme(legend.position = "bottom",
        axis.text.x = element_blank(), axis.ticks.x = element_blank())
```

Under hard bounds, 40.7\% of nodes have $\eta$ pinned at either
$\eta_\text{min}$ or $\eta_\text{max}$, creating a bimodal distribution
(skew $= -0.31$) of "consolidated" nodes (low $\eta$, high activity,
strong weights) and "exploratory" nodes (high $\eta$, low activity, weak
weights). Different assemblages accumulate different proportions of each
type, driving weight vector divergence. Under soft bounds, no nodes reach
the limits (by construction), $\eta$ is continuously distributed
(skew $= +0.55$, std $= 0.040$ vs $0.077$), and the reduced separation
yields lower CCD.

This extends a well-established principle in synaptic plasticity:
@vanrossum2000stable showed that additive (hard-bounded) STDP produces
bimodal weight distributions, while multiplicative (soft) rules produce
unimodal distributions. @miller1994role demonstrated that subtractive
constraints drive variables to extrema while multiplicative constraints
yield graded profiles. @fusi2007limits established that hard-bounded
synapses exhibit qualitatively different memory dynamics than soft-bounded
ones. Our contribution is showing that this same principle, applied one
level up---to the metaplastic learning rate rather than synaptic
weights---controls the strength of inter-assemblage differentiation.

A confound should be noted: under soft $\eta$ bounds, weights reach
$w_\text{max} = 1.0$ more frequently (mean maximum weight $= 1.000$) than
under hard $\eta$ bounds ($0.878$). This occurs because soft-bounded
$\eta$ values remain higher on average (not pinned to the floor), sustaining
stronger Hebbian updates that drive weights to the hard weight ceiling. The
weight saturation could independently reduce CCD by compressing the weight
space in which assemblages can diverge. Disentangling the $\eta$
discretisation effect from this weight saturation interaction remains for
future work.

Critically, however, local homeostasis under soft bounds still produces
statistically significant functional differentiation: response diversity
$= 0.015$ [0.012, 0.017] versus baseline $= 0.000$ [0.000, 0.000]
(Mann-Whitney $p < 10^{-40}$, Cohen's $d = 1.97$). The effect is
quantitatively weaker (32\% of the hard-bound value) but qualitatively
present. Local homeostasis produces genuine functional differentiation under
both bounding regimes; hard bounds amplify it through a bimodal
concentration effect that has well-characterised precedents in the synaptic
plasticity literature.

### Homeostatic timescale robustness

The EMA smoothing parameter controls the timescale of homeostatic sensing
(Exp 18). Fast sensing (EMA rate $= 0.10$) and slow sensing ($0.01$) both
produce strong differentiation: CCD $= 0.888$ and $0.869$ respectively,
compared to the default ($0.05$, CCD $= 0.841$). Response diversity is
similarly robust ($0.057$, $0.062$, $0.047$). The only notable difference
is assemblage count: slow sensing produces more assemblages ($11.1$ vs
$5.2$), likely because the sluggish target allows transient fluctuations
to seed additional individuation events before the homeostatic signal
catches up. The core differentiation result is invariant to homeostatic
timescale across an order of magnitude.

### Transplant test: differentiation is context-dependent

Experiment 19 tests whether assemblage identity is intrinsic or
context-dependent by swapping the $\eta$ values and activity histories of
the two largest assemblages at session 100. If differentiation reflects
intrinsic assemblage properties (DeLanda's relations of exteriority),
transplanted assemblages should retain their functional profiles in their
new spatial context.

They do not. The transplant produces no measurable disruption at any
timescale: CCD, response diversity, and assemblage count are
indistinguishable from the control condition by session 101---one session
(300 steps) after the swap. The transplanted $\eta$ values are immediately
overwritten by the local homeostatic target at the new spatial location.
Assemblage identity is determined entirely by spatial context, not by
intrinsic properties of the component nodes.

This result bears on DeLanda's [-@delanda2006new] relations of exteriority,
though the falsification is not absolute. DeLanda acknowledges that
exteriority admits of degrees---components need not be fully detachable
to possess some degree of external relations. The assemblages do resist
perturbation (Exp 9), which indicates partial exteriority: they are not
infinitely context-sensitive. But the transplant result shows that the
*identity-constituting* relations are spatial, not intrinsic. The
differentiation is genuine (assemblages have measurably different
functional profiles) but not *portable*: it is constituted by the
relationship between assemblage and milieu, not by properties the
assemblage carries with it.

The self-sealing dynamics observed here---where consolidated pathways
preferentially reinforce themselves, resist perturbation, and maintain
their organisation through their own activity---exhibit structural
parallels to operational closure as described in autopoietic theory
[@maturana1980autopoiesis]. Three levels of closure should be
distinguished. (a) *Simondonian operational closure*: the operation that
constituted the individual (Hebbian reinforcement of co-active edges) is
the same operation that now maintains it. The assemblages satisfy this
criterion. (b) *Varela's organisational closure*: the network of processes
that produces the system's components is identical to the network that
those components constitute. This is stronger --- it requires the system
to produce its own processes, not just maintain its own state. The
assemblages do not satisfy this criterion: the Hebbian rule is imposed,
not self-produced. (c) *Full autopoiesis*: organisational closure plus
self-production of a topological boundary in physical space
[@beer2020autopoiesis]. The assemblages do not produce their own boundary
--- it emerges from the weight dynamics but is not a distinct physical
structure. This positions the assemblages at level (a), between minimal
autonomy and full autopoiesis in the hierarchy described by Thompson
[-@thompson2007mind].

In Simondon's terms, the assemblage and its
associated milieu are co-individuated---neither has an identity independent
of the other. In DeLanda's terms, relations of exteriority do not hold:
components cannot be detached from their assemblage and plugged into another
while retaining their functional identity.


## Discussion

### The three-way comparison

The central finding of this paper can be stated as a three-way comparison
across the full programme:

| Homeostatic regime | Outcome |
|--------------------|---------|
| None (Papers 1--3) | Self-sealing: individuation without differenciation |
| Global target | Monopoly: one assemblage dominates |
| Local target | Stable differentiation: multiple distinct assemblages |

Self-sealing, previously diagnosed as the obstacle to differenciation (Paper
3), is reinterpreted here as its *precondition*. Without self-sealing, no
assemblage maintains its boundaries long enough to consolidate a distinct
identity. With self-sealing but no homeostatic regulation, assemblages
individuate but remain clones. With self-sealing and local homeostatic
regulation, each assemblage consolidates in its own neighbourhood, subject to
its own local reference signal, producing genuinely different internal
dynamics.

### Why Turing partially fails on Hebbian substrates

The Turing result is more nuanced than a simple falsification. Two structural
features of the Hebbian substrate are genuinely incompatible with the
reaction-diffusion framework, while a third turned out to be partly
artifactual:

1. **Dynamic topology.** Classical Turing analysis assumes a fixed diffusion
   graph. The Hebbian substrate rewires at every step---the Laplacian
   eigenspectrum is non-stationary. This is a structural incompatibility
   that no implementation change can resolve.

2. **Sparse connectivity.** @wolfrum2012turing showed that sparse networks
   generically produce single-differentiated-node solutions. Distributed
   Turing patterns require dense, regular connectivity.

3. **Absorbing $\eta$ boundaries (partly artifactual).** Under hard
   learning rate clipping ($\eta_i \in [\eta_\text{min}, \eta_\text{max}]$
   with hard clip), bounded $\eta$ values create absorbing states where
   inhibitor coupling cannot modulate the learning rate once it reaches a
   bound. Experiment 11 showed that soft $\eta$ bounding (tanh saturation)
   partially rescues coupling, confirming this was an implementation
   artifact rather than a fundamental incompatibility. This matters
   philosophically: the Deleuzian framework from which we derived the Turing
   hypothesis assumes continuous intensive quantities
   [@deleuze1994difference, ch. 5]. Hard $\eta$ clipping violates this
   assumption, and our initial failure to recognise this introduced an
   unnecessary discontinuity into the very mechanism we were testing.

The honest assessment is therefore: Turing morphogenesis on this substrate is
*structurally impaired* (points 1--2) and was *additionally sabotaged* by an
implementation choice inconsistent with its theoretical grounding (point 3).
With soft $\eta$ bounds, coupling achieves CCD $= 0.65$---genuine
differentiation, though still significantly below local homeostasis
(CCD $= 0.84$ under hard $\eta$ bounds, $0.30$ under soft). The structural
impairments explain the residual gap.

### Philosophical grounding: a productive fracture

This programme began within a Deleuzian framework: individuation as
actualisation of virtual multiplicities, differenciation as the production of
genuinely novel kinds through intensive processes
[@deleuze1994difference]. Papers 1--3 found that Hebbian dynamics produce
individuation (actualisation) but only bare repetition (same process,
identical copies). The question was: what produces complex repetition?

The answer---local homeostatic targets---creates tensions within this
framework that we believe are more productive to name than to resolve. A
third philosophical voice, DeLanda's assemblage theory, provides a diagnostic
that neither Deleuze nor Simondon alone supplies.

**Where Deleuze holds.** Local targets create genuinely heterogeneous
problematic fields in the Deleuzian sense. Each node confronts a *singular*
problem defined by its neighbourhood's contingent history. The
differentiation that emerges is not prescribed by the targets---the targets
co-evolve with the dynamics, and perturbation experiments confirm the
outcomes are not determined by specific parameter values. Different
localities constitute different differential fields, each demanding its own
resolution. This is differenciation through the problem
[@deleuze1994difference, ch. 4].

The soft-bounds finding reinforces this reading for the Turing mechanism.
Our initial Turing test violated a Deleuzian premise: intensive quantities
are continuous, and hard $\eta$ clipping introduced a non-Deleuzian
discontinuity. When we restored continuity (soft $\eta$ bounds), the
Deleuzian mechanism partially succeeded (CCD $= 0.65$).

However, the bound-type factorial (Exp 13) reveals a tension that must be
confronted honestly. The strongest local homeostasis result (CCD $= 0.84$)
depends on the same hard $\eta$ bounds that we criticise for sabotaging the
Turing mechanism. Under soft $\eta$ bounds, local homeostasis produces
CCD $= 0.30$---weaker geometric separation, with response diversity of $0.015$ confirming genuine but reduced functional differentiation. The philosophical
framework cannot simultaneously condemn hard $\eta$ bounds for Turing and
rely on them for homeostasis without addressing the asymmetry.

The resolution lies in Simondon's distinction between *moulding* and
*modulation* [@simondon2020individuation]. Moulding imposes form from
without through fixed boundary conditions; modulation continuously regulates
an ongoing process through variable constraints. Hard $\eta$ bounds create
absorbing states that classify nodes into binary types (consolidated vs.
exploratory)---this is moulding. The discretisation is not intrinsic to
the dynamics but imposed by the boundary. Soft $\eta$ bounds preserve the
intensive character of the learning rate, allowing continuous
regulation---this is modulation. The stronger differentiation under hard
bounds is Simondonian moulding: effective but externally constrained.
Moulding is not pejorative in Simondon's framework---it is a legitimate
operation, as in casting, that forecloses further individuation while
producing stable form. The weaker differentiation under soft bounds is
Simondonian modulation: open-ended but dynamically modest. Moulding is,
in Simondon's analysis, a limit case of modulation---modulation frozen at
its boundary values.

Both regimes produce functional differentiation (response diversity $> 0$).
The moulding regime amplifies it through a discretisation effect that has
well-characterised precedents in the synaptic plasticity literature
[@vanrossum2000stable; @miller1994role; @fusi2005cascade]. Whether
biological metaplastic variables exhibit hard or soft saturation is an
empirical question; evidence supports both possibilities
[@abraham2001heterosynaptic; @he2016experience].

**Where Simondon wins.** The critical period finding introduces an
irreversible temporal ordering: individuation must precede communication.
Coupling before consolidation destroys differentiation; coupling after
preserves it. The outcomes are qualitatively different and the ordering
cannot be reversed. This is Simondon's transduction---the coupled
individuation of milieu and individual, where the milieu individuates
alongside the entity it supports [@simondon2020individuation].

Deleuze's treatment of time in *Difference and Repetition* (Chapter 2) is
not achronological---the three syntheses of time (habit, memory, eternal
return) provide a rich temporal ontology. But Deleuze resists *irreversible
developmental ordering*: the eternal return as third synthesis implies that
all configurations of the virtual are in principle revisitable. The critical
period contradicts this. Once assemblages consolidate under local
homeostatic targets, the basin structure is carved; introducing coupling
*before* this consolidation produces a qualitatively different (and
irrecoverable) outcome. This irreversibility maps naturally onto Simondon's
transduction, which is explicitly directional, and sits uneasily with
Deleuze's emphasis on the virtual as eternally available for re-actualisation.

**Where DeLanda diagnoses.** The functional probe (Exp 12) operationalises
DeLanda's criterion for genuine capacity differentiation
[@delanda2006new; @delanda2002intensive]. DeLanda distinguishes between
properties (actual, denumerable) and capacities (virtual, forming an open
list). Genuine differentiation requires that components possess qualitatively
different *sets of capacities*, not merely different positions or parameter
values.

The baseline result is the critical test: CCD $= 0.90$ but response
diversity $= 0.000$. Assemblages that are geometrically distinct are
functionally identical---different instances of a single type, not different
kinds. DeLanda's five conditions for capacity formation
[@delanda2006new, ch. 1] map onto our substrate as follows:

| DeLanda condition | Status |
|---|---|
| 1. Heterogeneous components | Yes: $\eta$ divergence $> 0$, response rank $> 1$ (Exps 12, 17) |
| 2. Relations of exteriority | **No**: transplant shows identity is context-dependent (Exp 19) |
| 3. Intensive processes + thresholds | Partial: gradients present, no kind-transition |
| 4. Semi-permeable boundaries | Replaced by developmental sequence (Exp 8) |
| 5. Top-down constraint preserving variation | Yes: local targets (Exp 9 perturbation resilience) |

Three of five conditions are satisfied, one is temporally substituted, and
one---relations of exteriority---is experimentally falsified. The
transplant result (Exp 19) shows that assemblage identity cannot be detached
from spatial context: components relocated to a new neighbourhood are
immediately assimilated, their transplanted $\eta$ values overwritten by
the local homeostatic target within a single session. This is not attractor
stability (which Exp 9 confirms) but its opposite: the attractor is
spatially indexed, and nodes converge to whichever basin their spatial
position defines, regardless of their initial state.

This falsification is itself a contribution. It precisely characterises the
kind of differentiation this substrate achieves: Simondonian
co-individuation (assemblage and milieu constitute each other) rather than
DeLandian exteriority (components retain portable identity). The
differentiation is real but relational---it exists in the coupling between
entity and context, not in the entity alone.

**The fracture.** The system may achieve *differentiation*---symmetry-breaking
into basins with measurably different functional profiles---without achieving
*differenciation* in the full Deleuzian sense---the production of
qualitatively novel virtual structures. The vocabulary of outcomes (activity
level, connectivity, learning rate, gain) is fixed from the start; the
system selects among pre-given possibilities rather than creating new ones.
Even the functional differentiation documented in Exp 12 is parametric:
assemblages vary in the *value* of their signal amplification gain, not in
the *kind* of function they perform. Whether parametric variation across a
fixed capacity space constitutes genuine novelty or mere distribution is an
open philosophical question that computational evidence alone cannot
adjudicate.

We propose that this fracture between differentiation and differenciation
marks a precise boundary: what a purely Hebbian substrate with homeostatic
regulation *can* and *cannot* do. It can break symmetry, consolidate distinct
attractors, resist perturbation, and produce assemblages with measurably
different functional profiles. It cannot generate qualitatively new
dynamics that were not latent in its initial rule-set. If differenciation in
the strong Deleuzian sense requires the production of genuinely novel
capacities---new kinds of rules, not new values of existing rules---then even
the locally regulated Hebbian substrate falls short. But it falls short in a
precisely characterised way that points toward what a next-generation
substrate would require: rule-space expansion, not merely parameter-space
exploration. Adams et al. [-@adams2017formal] demonstrated formally that
only state-dependent update rules---where the system's current state
modifies the rule governing its next state---produce unbounded evolutionary
dynamics. A fixed Hebbian rule, however richly parameterised, constrains
the system to a bounded transition graph that must eventually repeat. The
metaplastic extension localises a *parameter* ($\eta$) but not the *form*
of the rule (Hebbian correlation). Crossing the productive fracture
requires making the update equation itself state-dependent---a direction
exemplified by Flow-Lenia's parameter localisation
[@plantec2023flowlenia].


## Conclusion

Assemblages under local homeostatic regulation develop genuinely independent
functional profiles: response rank reaches $2.4$ [$1.7$, $3.0$] under
the standard pruning regime (Exp 17) and $5.6$ when pruning is removed
(Exp 17np; see Paper 4S), compared to a baseline of $1.2$ (parametric).
Response diversity increases from $0.000$ (baseline) to $0.047$ (local
homeostasis), and $\eta$ divergence between assemblages reaches $F = 4.9$
($p < 0.0001$). These independent functional probes establish that local
homeostatic targets produce genuine capacity differentiation, not merely
geometric separation (which CCD alone cannot distinguish; Paper 4S). The
resulting differentiation is sustained (1000 sessions),
perturbation-resilient (50\% scramble, full recovery), robust to structural
parameters (grid size, locality, target radius), and exhibits critical period
dynamics (coupling after consolidation preserves diversity; coupling before
consolidation destroys it).

Four findings carry forward. First, the partial Turing rescue under soft
$\eta$ bounds reveals that the original failure was partly an implementation
artifact inconsistent with the Deleuzian framework that motivated the
hypothesis---a methodological lesson in attending to a theory's own
assumptions when designing its tests. Second, the functional probe
establishes that geometric differentiation (distinct weight structures) can
exist without functional differentiation (distinct capacities), vindicating
DeLanda's insistence that genuine differentiation requires capacity
divergence, not merely spatial separation. Third, the bound-type sensitivity
is itself a finding: hard $\eta$ bounds amplify differentiation through a
discretisation effect (Simondonian moulding) that extends a well-known
principle from synaptic to metaplastic dynamics
[@vanrossum2000stable; @miller1994role]. Soft $\eta$ bounds produce weaker
but philosophically cleaner differentiation (Simondonian modulation). Both
regimes produce genuine functional differentiation; the bound type controls
its strength. Fourth, the differentiation achieved---while real, stable,
and functional---remains parametric: different values of the same
capacities, not qualitatively novel kinds. If complex repetition in the full
Deleuzian sense requires the production of genuinely new capacities, then
differenciation in Hebbian substrates requires not merely heterogeneous
regulation but *rule-space expansion*---a substrate capable of generating new
kinds of dynamics, not just distributing existing ones across space.

What would differenciation look like computationally? We can now state this
precisely: assemblages that develop qualitatively different *update rules*,
not just different parameter values of the same rule. In our substrate, all
nodes execute the same Hebbian learning with homeostatic $\eta$ adjustment;
the rules are parametrically varied ($\eta$ values differ) but structurally
identical. Genuine differenciation would require some assemblages to
spontaneously develop anti-Hebbian dynamics, competitive inhibition, or
temporal integration properties not present in the initial rule-set. This
connects to the open-ended evolution literature's distinction between
variation within a fixed genotype-phenotype map and the expansion of the map
itself [@aguilar2014openended; @banzhaf2016defining]. The transplant
result (Exp 19) sharpens the point: even the parametric variation achieved
is context-dependent, immediately overwritten when the spatial context
changes. Portable, intrinsic functional identity---the hallmark of
DeLandian exteriority and a prerequisite for open-ended
evolution---requires a substrate whose components can carry their rules
with them. Flow-Lenia's parameter localisation [@plantec2023flow]
achieves this; our Hebbian substrate does not. The gap between local
homeostatic differentiation and genuine rule-carrying individuation defines
the research frontier for the next phase of this programme.


## Data and Code Availability

All simulation code, experiment runners, and analysis scripts are available
at [repository URL]. Raw parquet data files (approximately 260 MB across 18
experiments) are archived at [data repository URL]. All experiments use
fixed random seeds (0--19) for reproducibility. The simulation was executed
using Python 3.12.3, NumPy 1.26, PyArrow 14.0, scikit-learn 1.3, and Ray
2.54.0 on a three-node cluster (68 CPUs, 2 GPUs). Figures were generated
using R 4.3 with ggplot2, patchwork, and ggthemes via Quarto.


## References
